\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage{float}

\title{Pipeline Dataset Brief}
\author{}
\date{}

\begin{document}
\maketitle

\noindent\textbf{File:} \texttt{PIPELINE\_CAA\_00\_COMPLETE.csv} \\
\textbf{Dimensions:} 66,655 rows, 35 columns \\
\textbf{Source:} EPA ECHO --- CAA Compliance Pipeline \\
\textbf{Documentation:} \url{https://echo.epa.gov/tools/data-downloads/caa-pipeline-download-summary}

\section{Unit of Observation}

Each row is a \textbf{violation $\times$ enforcement action $\times$ evaluation} combination. A single violation can appear in multiple rows if it is linked to multiple enforcement actions or evaluations. The unique key is \texttt{SOURCE\_ID} + \texttt{VIOL\_ACTIVITY\_ID} + \texttt{EA\_ACTIVITY\_ID} + \texttt{EVAL\_ACTIVITY\_ID}.

\begin{itemize}[nosep]
  \item 52,346 distinct violations (\texttt{VIOL\_ACTIVITY\_ID})
  \item 20,251 distinct facilities (\texttt{SOURCE\_ID}, equivalent to \texttt{PGM\_SYS\_ID} in ICIS-Air)
  \item 78\% of violations appear in exactly 1 row; 16\% in 2 rows; 4\% in 3 rows; $<$1\% in 4+
\end{itemize}

\section{Data Generating Process}

Per EPA documentation, the ECHO pipeline ``shows links between Compliance Monitoring Activities (CMA) to any related violations and/or enforcement actions.'' \texttt{FOUND\_VIOLATION = Y} for all 66,655 rows. However, 10.8\% of rows carry system-generated \texttt{VIOL\_ACTIVITY\_ID}s rather than real violations (see below). Per EPA, these IDs ``did not have an actual violation activity identification number'' and were ``system generated for purposes of creating the pipeline table.'' Facilities with clean compliance records do not appear.

\subsection*{PIPELINE\_FLAG}

Per EPA documentation: ``An internally generated flag to indicate at least [one] violation (\texttt{VIOL\_ACTIVITY\_ID}) linked to an evaluation and/or enforcement action.'' When set to \texttt{Y}, the pipeline table is displayed on the facility's ECHO Detailed Facility Report.

\begin{table}[H]
\centering
\begin{tabular}{llrrrr}
\toprule
Flag & Rows & Facilities & Has EA & Has Eval \\
\midrule
Y & 58,251 (87.5\%) & 15,792 & 78\% & 52.4\% \\
N & 8,404 (12.5\%) & 4,459 & 0\% & 0\% \\
\bottomrule
\end{tabular}
\end{table}

\noindent\textbf{Empirical pattern:} All 8,404 rows with \texttt{PIPELINE\_FLAG = N} have \texttt{EA\_FLAG = N} and \texttt{EVAL\_FLAG = N}---no linked enforcement actions or evaluations. These rows retain violation-level fields (type, dates). It is not clear from the documentation whether N rows once had linkages that were removed, or whether they never had linkages.

\subsection*{Placeholder rows (no actual violation)}

Not every row represents a real violation. 7,193 rows (10.8\%) carry \textbf{system-generated} \texttt{VIOL\_ACTIVITY\_ID}s---the EPA documentation states that values starting with \texttt{9906} or \texttt{9913} ``did not have an actual violation activity identification number'' and were ``system generated for purposes of creating the pipeline table.'' These fall into two categories:

\begin{enumerate}[nosep]
  \item \textbf{\texttt{VIOL\_TYPE = NA}} (4,067 rows, IDs starting with \texttt{9906}): These rows have \texttt{EA\_FLAG = Y} but no violation dates or type.

  \item \textbf{\texttt{VIOL\_TYPE = ``Linked to Viol.\ Below''}} (3,126 rows, IDs starting with \texttt{9913}): These rows have \texttt{EA\_FLAG = N}, \texttt{EVAL\_FLAG = N}, and no violation dates.
\end{enumerate}

\noindent The remaining 59,462 rows (89.2\%) carry real violation IDs (10-digit, starting with \texttt{3400} or \texttt{3600}) and a substantive \texttt{VIOL\_TYPE} of HPV or FRV. \textbf{For analytical use, filter to \texttt{VIOL\_TYPE \%in\% c("HPV", "FRV")} to retain only real violations.}

\subsection*{``Date Out of Range'' enforcement actions}

\texttt{EA\_TYPE = ``Date Out of Range''} appears in 1,116 rows (2.5\% of EA-linked rows). All 1,116 have no \texttt{EA\_DATE} recorded. The cause is not documented by EPA. These rows should be treated as having missing enforcement timing.

\section{Summary Statistics}

\subsection*{Violations}

Filtering to real violations (\texttt{VIOL\_TYPE} $\in$ \{HPV, FRV\}), the pipeline contains 45,164 distinct violations across 59,462 rows. FRV violations outnumber HPV roughly 2.4:1 (42,037 rows vs.\ 17,425 rows). Of the 45,164 distinct real violations, 81.4\% have \texttt{PIPELINE\_FLAG = Y} and 18.6\% have \texttt{PIPELINE\_FLAG = N}.

Among resolved violations with valid dates ($n$ = 9,222), the median time to resolution is \textbf{180 days} (mean 304 days; IQR 56--384 days).

\subsection*{Enforcement actions}

68\% of rows (45,340) have an enforcement action linked (\texttt{EA\_FLAG = Y}). The distribution of EA types:

\begin{table}[H]
\centering
\begin{tabular}{lrr}
\toprule
EA Type & Rows & \% of EA rows \\
\midrule
Notice of Violation & 23,818 & 52.5\% \\
Administrative - Formal & 16,782 & 37.0\% \\
Warning Letter & 2,318 & 5.1\% \\
Date Out of Range & 1,116 & 2.5\% \\
Judicial & 380 & 0.8\% \\
Other / State-local & 926 & 2.0\% \\
\bottomrule
\end{tabular}
\end{table}

\subsection*{Penalties and compliance action costs}

14,427 rows carry a non-zero penalty (\texttt{EA\_PENALTY\_AMT}), totaling \$1.93 billion. The distribution is heavily right-skewed: median \$7,350, mean \$133,818, max \$47.8M.

Compliance action costs (\texttt{EA\_COMP\_ACTION\_COST}) are sparse---only 94 rows have a non-zero value---but large in magnitude, totaling \$849M.

\subsection*{Lead agency}

\texttt{VIOL\_LEAD\_AGENCY} identifies which agency leads the violation case. Recoding the 74 raw values into three categories:

\begin{table}[H]
\centering
\begin{tabular}{lrrrrr}
\toprule
Agency & Rows & Facilities & \% with EA & Total Penalties & Median Penalty \\
\midrule
State & 59,262 (88.9\%) & 18,812 & 69.5\% & \$1.61B & \$6,750 \\
EPA & 194 (0.3\%) & 114 & 74.7\% & \$3.5M & \$75,000 \\
Local & 6 ($<$0.1\%) & 4 & 100\% & \$964K & \$482K \\
\bottomrule
\end{tabular}
\end{table}

\noindent The remaining 7,193 rows (10.8\%) have no lead agency---these are the placeholder rows with system-generated \texttt{VIOL\_ACTIVITY\_ID}s. State agencies lead the vast majority of cases. EPA-led cases are rare but carry substantially higher median penalties (\$75,000 vs.\ \$6,750 for state-led).

\section{Example Rows}

\subsection*{Example A --- Unresolved FRV, no enforcement action yet}

\begin{table}[H]
\centering
\begin{tabular}{ll}
\toprule
Field & Value \\
\midrule
\texttt{SOURCE\_ID} & OK0000004005100121 \\
\texttt{AIR\_NAME} & DCP Operating Co LP / Wild Hog Booster Station \\
\texttt{PIPELINE\_FLAG} & Y \\
\texttt{EVAL\_FLAG / EVAL\_TYPE\_DESC} & Y / Self-Disclosure \\
\texttt{VIOL\_TYPE} & FRV \\
\texttt{VIOL\_START\_DATE} & 04/18/2024 \\
\texttt{VIOL\_END\_DATE} & --- \\
\texttt{EA\_FLAG} & N \\
\texttt{EA\_PENALTY\_AMT} & --- \\
\bottomrule
\end{tabular}
\end{table}

\noindent A facility self-disclosed a federally reportable violation. No enforcement action has been taken. Violation is still in the pipeline.

\subsection*{Example B --- HPV discovered by FCE, formal action with penalty}

\begin{table}[H]
\centering
\begin{tabular}{ll}
\toprule
Field & Value \\
\midrule
\texttt{SOURCE\_ID} & NVCCH00032003A0019 \\
\texttt{AIR\_NAME} & Titanium Metals Corp.\ Henderson Facility \\
\texttt{PIPELINE\_FLAG} & Y \\
\texttt{EVAL\_FLAG / EVAL\_TYPE\_DESC} & Y / FCE On-Site \\
\texttt{EVAL\_DATE} & 05/13/2024 \\
\texttt{VIOL\_TYPE} & HPV \\
\texttt{VIOL\_START\_DATE} & 08/05/2024 \\
\texttt{EA\_FLAG / EA\_TYPE} & Y / Administrative - Formal \\
\texttt{EA\_DATE} & 07/24/2024 \\
\texttt{EA\_PENALTY\_AMT} & \$11,150 \\
\bottomrule
\end{tabular}
\end{table}

\noindent An on-site FCE is linked to an HPV with a formal administrative action and an \$11,150 penalty. \texttt{PIPELINE\_FLAG = Y} despite the penalty being assessed.

\subsection*{Example C --- One violation, three rows (multiple EAs)}

\begin{table}[H]
\centering
\begin{tabular}{llll}
\toprule
\texttt{SOURCE\_ID} & \texttt{VIOL\_ACTIVITY\_ID} & \texttt{EA\_TYPE} & \texttt{EA\_DATE} \\
\midrule
MS0000002812100033 & 3400330404 & Notice of Violation & 10/09/2001 \\
MS0000002812100033 & 3400330404 & Date Out of Range & --- \\
MS0000002812100033 & 3400330404 & Date Out of Range & --- \\
\bottomrule
\end{tabular}
\end{table}

\noindent A single HPV at Holmes Group / Rival Division is linked to three enforcement action records. Two have \texttt{EA\_TYPE = ``Date Out of Range''} with no \texttt{EA\_DATE}. The violation appears in three rows because each EA linkage generates a separate row.

\subsection*{Example D --- Resolved violation}

\begin{table}[H]
\centering
\begin{tabular}{ll}
\toprule
Field & Value \\
\midrule
\texttt{SOURCE\_ID} & NC0000003703200251 \\
\texttt{AIR\_NAME} & North Carolina Central University \\
\texttt{PIPELINE\_FLAG} & N \\
\texttt{VIOL\_TYPE} & FRV \\
\texttt{VIOL\_START\_DATE} & 11/18/2021 \\
\texttt{VIOL\_END\_DATE} & 05/17/2023 \\
\texttt{EA\_FLAG} & N \\
\texttt{EVAL\_FLAG} & N \\
\bottomrule
\end{tabular}
\end{table}

\noindent \texttt{PIPELINE\_FLAG = N} with a violation start and end date spanning $\sim$18 months. Consistent with the pattern described above: \texttt{EA\_FLAG = N} and \texttt{EVAL\_FLAG = N}.

\section{Key Variables}

\textbf{Violation fields:}
\begin{itemize}[nosep]
  \item \texttt{VIOL\_TYPE}: HPV (high priority, 26\%) or FRV (federally reportable, 63\%). The remaining 11\% are placeholder rows: 6\% \texttt{NA} (system-generated IDs with no actual violation) and 5\% ``Linked to Viol.\ Below.''
  \item \texttt{VIOL\_ACTIVITY\_ID}: the key linking violations to CMAs and enforcement actions. IDs starting with \texttt{9906} or \texttt{9913} are system-generated placeholders (per EPA documentation).
  \item \texttt{VIOL\_START\_DATE}: when the violation was determined; missing for 10.8\% of rows (all placeholder rows lack dates).
  \item \texttt{VIOL\_END\_DATE\_DATE}: violation end date; missing for 74.6\% of rows.
  \item \texttt{VIOL\_PROGRAMS}: which CAA programs were violated (CAATVP, CAASIP, CAAMACT, etc.).
\end{itemize}

\textbf{Enforcement action fields} (only for \texttt{EA\_FLAG = Y}, 68\% of rows):
\begin{itemize}[nosep]
  \item \texttt{EA\_TYPE}: Notice of Violation (52\%), Administrative - Formal (37\%), Warning Letter (5\%), Date Out of Range (2.5\%), Judicial (1\%).
  \item \texttt{EA\_DATE}: date of the enforcement action. Missing for all ``Date Out of Range'' rows.
  \item \texttt{EA\_PENALTY\_AMT}: monetary penalty; non-zero for 14,427 rows. Median \$7,350, mean \$133,818, max \$47.8M.
  \item \texttt{EA\_COMP\_ACTION\_COST}: supplemental environmental project cost; non-zero for only 94 rows.
\end{itemize}

\textbf{Evaluation fields} (only for \texttt{EVAL\_FLAG = Y}, 46\% of rows):
\begin{itemize}[nosep]
  \item \texttt{EVAL\_TYPE\_DESC}: PCE Off-Site (26\%), FCE On-Site (15\%), Investigation (11\%), Self-Disclosure (11\%).
  \item \texttt{EVAL\_LEAD\_AGENCY}: State (84\%), EPA (8\%), Local (8\%).
  \item \texttt{EVAL\_ACTIVITY\_ID}: set to \texttt{-9999} when no evaluation is linked (43,084 rows, 65\%).
\end{itemize}

\textbf{Flags and metadata:}
\begin{itemize}[nosep]
  \item \texttt{PIPELINE\_FLAG}: per EPA, ``an internally generated flag to indicate at least [one] violation linked to an evaluation and/or enforcement action.'' In the data, all N rows have \texttt{EA\_FLAG = N} and \texttt{EVAL\_FLAG = N}.
  \item \texttt{EVAL\_FLAG}: per EPA, ``an internally generated flag to indicate if there is at least one Compliance Monitoring Activity (CMA) (evaluation) to the violation activity id.''
  \item \texttt{EA\_FLAG}: per EPA, ``a system generated flag indicating if there is an enforcement action.''
  \item \texttt{VIOL\_FLAG}: per EPA, ``a system generated flag indicating a violation is present. Entries noted as violation not found will have a generated \texttt{VIOL\_ACTIVITY\_ID} to use as links between Evaluations and Enforcement Actions.''
  \item \texttt{OFFICIAL\_FLAG}: per EPA, ``this flag is set to `Y' when an entry is counted as compliance monitoring strategy activities.''
  \item \texttt{FEA\_ISSUE\_DATE\_FLAG}: Y = a formal enforcement action issue date exists (224 rows).
  \item \texttt{SORT\_ORDER}, \texttt{SORT\_DATE}: system-generated fields to aid linkage creation and display ordering.
\end{itemize}

\section{Temporal Coverage}

\begin{itemize}[nosep]
  \item Violation start dates span 1994--2026, but 95\%+ are 2016--2025
  \item Median enforcement lag (EA date $-$ violation start): 42 days
  \item 25.4\% of enforcement actions are same-day as the violation (lag = 0)
  \item 3,844 cases where EA date precedes violation start date (negative lag; cause undocumented)
\end{itemize}

\section{Join Keys to Other Datasets}

\begin{table}[H]
\centering
\begin{tabular}{lll}
\toprule
Pipeline field & ICIS-Air equivalent & Join rate \\
\midrule
\texttt{SOURCE\_ID} & \texttt{PGM\_SYS\_ID} & 100\% (20,251/20,251) \\
\texttt{REGISTRY\_ID} & \texttt{REGISTRY\_ID} (FRS) & 99.9\% (19,672/19,687) \\
\texttt{VIOL\_ACTIVITY\_ID} & \texttt{ACTIVITY\_ID} (VIOLATION\_HISTORY) & 86.3\% (45,164/52,346) \\
\texttt{EA\_ACTIVITY\_ID} & \texttt{ACTIVITY\_ID} (FORMAL/INFORMAL) & 98.9\% (40,024/40,478) \\
\texttt{EVAL\_ACTIVITY\_ID} & \texttt{ACTIVITY\_ID} (FCES\_PCES) & 66.4\% (12,199/18,360) \\
\bottomrule
\end{tabular}
\end{table}

\noindent 7,182 pipeline violations are \emph{not} in the ICIS-Air violation history table---these are the system-generated placeholder IDs. Conversely, 55,983 violation history records are \emph{not} in the pipeline. The pipeline is not a superset of all violations.

\section{What the Pipeline Adds Over Individual ICIS-Air Tables}

The individual ICIS-Air tables share \texttt{ACTIVITY\_ID} and \texttt{PGM\_SYS\_ID} but do not directly link violations to enforcement actions or evaluations. The pipeline pre-joins these linkages. Unique to the pipeline:

\begin{enumerate}[nosep]
  \item \textbf{Violation-to-enforcement linkage} --- which EA is linked to which violation
  \item \textbf{Violation resolution status} --- \texttt{PIPELINE\_FLAG} and \texttt{VIOL\_END\_DATE}
  \item \textbf{Time-to-resolution} --- from \texttt{VIOL\_START\_DATE} to \texttt{VIOL\_END\_DATE\_DATE}
  \item \textbf{Supplemental compliance costs} --- \texttt{EA\_COMP\_ACTION\_COST} (not in formal actions table)
\end{enumerate}

\section{Potential Use in a Facility-Year Panel}

To incorporate the pipeline into a facility-year panel, the data would need to be collapsed in two steps: (1) deduplicate to the violation level by grouping on \texttt{SOURCE\_ID} + \texttt{VIOL\_ACTIVITY\_ID}, and (2) aggregate to facility-year using the violation start date. After filtering to real violations (\texttt{VIOL\_TYPE} $\in$ \{HPV, FRV\}), the pipeline could provide:

\begin{itemize}[nosep]
  \item \texttt{pct\_violations\_flag\_n} --- share of violations with \texttt{PIPELINE\_FLAG = N}
  \item \texttt{mean\_days\_to\_resolution} --- from \texttt{VIOL\_START\_DATE} to \texttt{VIOL\_END\_DATE\_DATE}
  \item \texttt{pct\_violations\_with\_ea} --- share of violations that received any enforcement action
  \item \texttt{pct\_hpv} --- share of violations classified as high priority vs.\ federally reportable
\end{itemize}

\noindent These variables are not constructable from the individual ICIS-Air tables, which lack the violation-to-EA and violation-to-evaluation linkages.

\textbf{Caveat:} The pipeline only contains facilities with violations. Compliant facilities would have no pipeline rows and would need to be coded as zeros or missing for these variables.

\end{document}
