\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{array}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{xcolor}

\titleformat{\section}{\large\bfseries}{}{0em}{}
\titlespacing*{\section}{0pt}{1.2em}{0.5em}
\setlist[itemize]{leftmargin=1.4em, itemsep=0.2em, topsep=0.3em}

\newcommand{\slidetag}[1]{\textcolor{gray!70}{\small\textsc{#1}}}

\title{\vspace{-2em}\textbf{Using Claude for Federal Environmental Data: Documentation \& Reproducibility}}
\author{}
\date{}

\begin{document}
\maketitle
\vspace{-3em}

%==================================================================
\section*{Problem \& Why}

EPA's Clean Air Act compliance data involves four disconnected systems: legacy
AFS, modern ICIS-Air, the enforcement ``pipeline,'' and the FRS facility
registry. Each has its own undocumented ID scheme, placeholder conventions, and
structural missingness. None of it is research-usable until it is documented,
validated, and linked across systems, and a single misread flag or unverified
ID match could corrupt everything downstream. Done by hand, the process of documenting and summarizing these datasets can be time-consuming and
monotonous.

%==================================================================
\section*{The Workflow}

I built the data documentation and summary tables with Claude, keeping the
entire process reproducible. Over the course of a few days, it ran the analysis scripts across 17 datasets and
more than 30 million rows, read the real output, and reported only what was
present in the data, with every figure and table traceable back to a script.

\begin{itemize}
  \item \textbf{Data dictionaries \& summary tables.} Claude wrote and ran the R scripts
        that profiled each dataset, its variable distributions, value counts,
        and missingness, then drafted data dictionaries and summary tables directly from
        that output.
  \item \textbf{Structural diagnosis.} It found that much of the missingness was
        not bad data but system artifacts like placeholder rows, sentinel
        $-9999$s, and flag conventions, and recorded those findings in the
        documentation rather than treating the gaps as errors.
  \item \textbf{Worked to my reproducibility standards.} The raw data was never
        modified, every derived file was rebuilt from code, and when a result
        shifted unexpectedly it was flagged rather than quietly smoothed over.
\end{itemize}

%==================================================================
\section*{Outcome \& Efficiency}

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}p{0.40\textwidth} p{0.28\textwidth} p{0.22\textwidth}@{}}
\toprule
\textbf{Task} & \textbf{By hand} & \textbf{With Claude} \\
\midrule
Document and summarize 17 federal datasets (dictionaries + tables)
  & $\sim$2--3 weeks & $\sim$1--2 days \\
Structural missingness diagnosis
  & $\sim$3--5 days & $\sim$1 hour \\
Empirical-only audit of every documented claim
  & $\sim$1--2 days per dataset & minutes \\
\midrule
\textbf{Total}
  & \textbf{$\sim$4--5 weeks of solo work}
  & \textbf{$\sim$1 week of sessions} \\
\bottomrule
\end{tabular}
\end{center}

\section*{What That Time Bought}

\begin{itemize}
  \item \textbf{Cross-dataset validation.} Reconciled record counts and variable definitions across the separate systems, catching mismatches that would otherwise surface much later.
  \item \textbf{Reproducible publishing.} Set up a pipeline that pushes the
        summary tables directly to shared Google Sheets / Drive, so
        collaborators work from live outputs rather than emailed spreadsheets.
  \item \textbf{Policy context \& literature.} Time to engage with the
        regulatory background, how CAA enforcement and emissions programs
        operate, and the literature on environmental monitoring and enforcement.
\end{itemize}

\end{document}
